Hybrid Threads for the Parrot Virtual Machine

### 2012-02-23 Paper.tex
First round of fixes

On July 19th 2000, the Perl 6 design process was announced. Perl 5 had been a very flexible and widely used programming language but had started to show it's age and suffered from early design decisions. The Perl interpreter is written in C and has accumulated a lot of cruft over the years. The general consensus among the core developers was that the code had reached a state where maintenance was approaching impossibility. [http://use.perl.org/~masak/journal/40451]. An attempt to reimplement these internals had failed but led to the decision that the interpreter for a Perl 6 language should be developed independently of the needs of Perl 5. Since the Perl 6 syntax was very much in flux (and parts of it still are) the designers of these new internals tried to work very independently of any syntax related questions. [http://www.developer.com/lang/perl/article.php/10940\_3076571\_3/Perl-6-and-the-Parrot-Project.htm]. Taking up the name born in an April Fool's Day joke announcing the mergin of the Perl and Python programming languages, these new internals were called Parrot. [http://www.perl.com/pub/2001/04/01/parrot.htm]

@@

The Perl 6 design process began with asking the Perl users for what they were expecting from the new version of the language. The very first feature that got asked for was well integrated multi-threading support [http://dev.perl.org/perl6/rfc/1.html]. Perl 5 had two different implementations of thread support. In the first model, called 5005threads, data was shared by default and shared access to data had to be explicitely synchronized. This was similiar to the models used by languages such as C or Java. The implementation however suffered from data corruption and crashes and thus was not recommended for production use [http://search.cpan.org/~nwclark/perl-5.8.8/lib/Thread.pm]. Perl 5.6 introduced the newer model called ithreads, mostly as a way to emulate fork() on Win32 platforms. Perl 5.8 exposed this API to the user of the programming language. In this new model, all data would is copied to each thread and afterwards thread local. Data must be explicitely shared between threads. In other words, in Perl threads are not lightweight at all. They have severe impact on memory usage, writes to shared data are expensive and still not all features of the language are usable in threaded programs.

@@

In a dynamic language, the object may change it's class at runtime. The inheritance hierarchy of the class may be changed by adding or removing parent classes. Methods may be added to or removed from classes (or objects) at runtime and even the way to find a method of a class may change. So a simple method call may result in the following steps:
%
\begin{itemize}
\item determining the class of the object
\item determining the method resolution method of the class
\item finding the actual method to call
\item calling the method
\end{itemize}
%
These steps have to be repeated for every following method call, because their results may change any time. In a threaded environment, a thread running in parallel may change the underlying data and meta data in between those sequences and even between those steps. As a consequence, this meta data has to be protected from corruption introducing the need for many locks in a very performance critical area.

@@

Before my work on this paper, Parrot did not have any threading support at all. The previous, defunct implementation had been removed.

@@

Control flow is modeled in Parrot using continuation passing. A continuation is a data structure containing the a state of a program at a given point in it's execution. It contains all information which is needed to continue a program in a certain state, e.g. a call stack, the instruction pointer and contents of local variables. ``Calling'' a continuation means restoring the state encapsulated in the continuation.

When a function is called, instead of pushing all this information on the stack, the function is given a return continuation as part of it's parameters. Returning from a function means calling the return continuation with possible return values stored in the registers where the calling code is expecting them. Continuation passing makes things like tail call optimization \footnote{When as the last statement of a function another function is called and it's value returned unchanged, the outer function may just pass on it's return continuation thus saving the need to create a new continuation and one step on returning from the nested function} very simple and as we will see later will be a very important piece in implementing green threads.

@@

In a more complicated example, the array could have to resize it's data buffer to accomodate the new value. It would read the current size, allocate a new buffer, copy the values to the new buffer and then destroying the old one. Again if two threads try to do this simultainously, one of them could still be copying data, while the other is already at the point where it destroys the old buffer. This would lead to the copying thread accessing freed memory.

If we bring garbage collection into the mix, the possibilities for corruption grow even more. At the point where the first thread needs to allocate the new buffer, the garbage collector could decide that it needs to run and clear some unused memory. It would have to traverse the object tree to mark alive objects. Between the mark and the sweep runs, the second thread could change the pointer to the buffer from the old one to the new one. So the old buffer would have been marked alive, while during the sweep run the new buffer would be in it's place, but not yet marked alive.

@@

In Parrot green threads are called \textit{Task}. Each task is assigned a fixed amount of execution time. On expiry a timer callback sets a flag which is checked for at execution of every branch operation. Since the interpreter's state is well defined at this point, it's internal consistency is guaranteed. The same goes for the garbage collector. Since task preemption is only done while executing user level code, the GC can do it's work undisturbed and without the need for measures like locking. Since user level code is allowed to disable the scheduler, it can be guaranteed to run undisturbed through critical sections.

@@

In keeping the analogy of the interpreter being a software CPU, multithreading is implemented by having a separate interpreter with it's own register set for each thread. When a user starts a new task, the scheduler first looks for an idle thread. If one can be found, the task is scheduled on the thread's interpreter. If none can be found, a new thread with a new interpreter is started. Parrot tries to optimize the number of utilized threads by creating at most one for each CPU core in the system. If more tasks are started than the maximum number of threads, the tasks are distributed evenly among the running interpreters. This is effectively an implementation of the N:M threading model.

@@

For threads to be able to communicate with their creators and other threads, they still need to write to shared data. This is where green threads come into play. Since green threads are very light weight, it is feasible for a thread to create a task just for updating a variable. This task would be scheduled on the interpreter owning this variable. To reduce latency, the task is flagged to run immediately. The data owning interpreter would preempt the currently running task and process the new write task. Put another way, the data owning interpreter is simply told what to write to it's variables, so other threads don't have to.

@@

In previous attempts to implement threading support in Parrot, the answer to this problem was to disable the garbage collector while multiple threads are active. Other VMs solve this problem by suspending all threads while the GC is running.

>>>>

On July 19th 2000, the Perl 6 design process was announced. Perl 5 had been a very flexible and widely used programming language but had started to show its age and suffered from early design decisions. The Perl interpreter is written in C and has accumulated a lot of cruft over the years. The general consensus among the core developers was that the code had reached a state where maintenance was approaching impossibility. [http://use.perl.org/~masak/journal/40451]. An attempt to reimplement these internals had failed but led to the decision that the interpreter for a Perl 6 language should be developed independently of the needs of Perl 5. Since the Perl 6 syntax was very much in flux (and parts of it still are) the designers of these new internals tried to work very independently of any syntax related questions. [http://www.developer.com/lang/perl/article.php/10940\_3076571\_3/Perl-6-and-the-Parrot-Project.htm]. Taking up the name born in an April Fool's Day joke announcing the merging of the Perl and Python programming languages, these new internals were called Parrot. [http://www.perl.com/pub/2001/04/01/parrot.htm]

@@

The Perl 6 design process began with asking the Perl users for what they were expecting from the new version of the language. The very first feature that got asked for was well integrated multi-threading support [http://dev.perl.org/perl6/rfc/1.html]. Perl 5 had two different implementations of thread support. In the first model, called 5005threads, data was shared by default and shared access to data had to be explicitely synchronized. This was similiar to the models used by languages such as C or Java. The implementation however suffered from data corruption and crashes and thus was not recommended for production use [http://search.cpan.org/~nwclark/perl-5.8.8/lib/Thread.pm]. Perl 5.6 introduced the newer model called ithreads, mostly as a way to emulate fork() on Win32 platforms. Perl 5.8 exposed this API to the user of the programming language. In this new model, all data is copied to each thread and afterwards thread local. Data must be explicitely shared between threads. In other words, in Perl threads are not lightweight at all. They have severe impact on memory usage, writes to shared data are expensive and still not all features of the language are usable in threaded programs.

@@

In a dynamic language, the object may change its class at runtime. The inheritance hierarchy of the class may be changed by adding or removing parent classes. Methods may be added to or removed from classes (or objects) at runtime and even the way to find a method of a class may change. So a simple method call may result in the following steps:
%
\begin{itemize}
\item determining the class of the object
\item determining the method resolution method of the class
\item finding the actual method to call
\item calling the method
\end{itemize}
%
These steps have to be repeated for every following method call, because their results may change any time. In a threaded environment, a thread running in parallel may change the underlying data and meta data in between those sequences and even between those steps. As a consequence, this meta data has to be protected from corruption introducing the need for many locks in a very performance critical area.

@@

Before the work on this paper, Parrot did not have any threading support at all. The previous, defunct implementation had been removed.

@@

Control flow is modeled in Parrot using continuation passing. A continuation is a data structure containing the a state of a program at a given point in its execution. It contains all information which is needed to continue a program in a certain state, e.g. a call stack, the instruction pointer and contents of local variables. ``Calling'' a continuation means restoring the state encapsulated in the continuation.

When a function is called, instead of pushing all this information on the stack, the function is given a return continuation as part of its parameters. Returning from a function means calling the return continuation with possible return values stored in the registers where the calling code is expecting them. Continuation passing makes things like tail call optimization \footnote{When as the last statement of a function another function is called and its value returned unchanged, the outer function may just pass on its return continuation thus saving the need to create a new continuation and one step on returning from the nested function} very simple and as we will see later will be a very important piece in implementing green threads.

@@

In a more complicated example, the array could have to resize its data buffer to accomodate the new value. It would read the current size, allocate a new buffer, copy the values to the new buffer and then destroying the old one. Again if two threads try to do this simultainously, one of them could still be copying data, while the other is already at the point where it destroys the old buffer. This would lead to the copying thread accessing freed memory.

If we bring garbage collection into the mix, the possibilities for corruption grow even more. At the point where the first thread needs to allocate the new buffer, the garbage collector could decide that it needs to run and clear some unused memory. It would have to traverse the object tree to mark alive objects. Between the mark and the sweep runs, the second thread could change the pointer to the buffer from the old one to the new one. So the old buffer would have been marked alive, while during the sweep run the new buffer would be in its place, but not yet marked alive.

@@

In Parrot green threads are called \textit{Task}. Each task is assigned a fixed amount of execution time. On expiry a timer callback sets a flag which is checked for at execution of every branch operation. Since the interpreter's state is well defined at this point, its internal consistency is guaranteed. The same goes for the garbage collector. Since task preemption is only done while executing user level code, the GC can do its work undisturbed and without the need for measures like locking. Since user level code is allowed to disable the scheduler, it can be guaranteed to run undisturbed through critical sections.

@@

In keeping the analogy of the interpreter being a software CPU, multithreading is implemented by having a separate interpreter with its own register set for each thread. When a user starts a new task, the scheduler first looks for an idle thread. If one can be found, the task is scheduled on the thread's interpreter. If none can be found, a new thread with a new interpreter is started. Parrot tries to optimize the number of utilized threads by creating at most one for each CPU core in the system. If more tasks are started than the maximum number of threads, the tasks are distributed evenly among the running interpreters. This is effectively an implementation of the N:M threading model.

@@

For threads to be able to communicate with their creators and other threads, they still need to write to shared data. This is where green threads come into play. Since green threads are very light weight, it is feasible for a thread to create a task just for updating a variable. This task would be scheduled on the interpreter owning this variable. To reduce latency, the task is flagged to run immediately. The data owning interpreter would preempt the currently running task and process the new write task. Put another way, the data owning interpreter is simply told what to write to its variables, so other threads don't have to.

@@

In previous attempts to implement threading support in Parrot, the answer to this problem was to disable the garbage collector while multiple threads are active. Other VMs solve this problem by suspending all threads while the GC is running. While multi threaded garbage collection algorithms exist, their implementation would have been to complex and brittle for this paper [http://doc.cat-v.org/inferno/concurrent\_gc/].
